A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-08 sister/sister.tex
push
put text in

\begin{abstract}
New abstract.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

  * a general introduction to automatic theorem proving \\

  * relation to AI
   * among the earliest AI goals
   * yet somewhat disconnected from mainstream AI
     (with some exceptions, e.g. Urban ..., MaSh)

  * followed by an intro to SMT and what it's used for \\
  * followed to why theories are important (more efficient reasoning) \\
  * followed to why (co)datatypes are important

  * both for proving
   * not finitely axiomatizable
  * and for model finding
   * Nunchaku

\section{%The Theory of
(Co)datatypes}
\label{sec:co-datatypes}

xxx

\section{The Decision Procedure} % for (Co)datatypes}
\label{sec:the-decision-procedure}

\section{Evaluation} % on Isabelle Problems}
\label{sec:evaluation}

\section{Conclusion}
\label{sec:conclusion}

\newpage

X

\newpage

@@

\section{Introduction}
\label{sec:introduction}

@@

%  * useful both for proving and for model finding

%\ref{sec:examples}

%  * benchmarking is often an issue -- esp. codatatypes
%
%Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall

%  * setting: FOL
%    * core procedure is restricted to ground
%      * but theory solver cooperates

%  * codatatypes were added later, motivated by the use of SMT solvers as
%    backends to proof assistants (more specifically, CVC4 to Isabelle/HOL)

%  * setting :
%    * universal formulas
%    * many-sorted logic
%   * mutually (co)recursive types with constructors, selectors, and
%      discriminators

%  * codatatypes: from a theoretical and implementational point of view, like
%      datatypes but:
%    * infinite values (infinitely many nested constructors)
%    * codatatypes are never empty (e.g. finite streams are rejected)

%\[
%      \keyw{codatatype}\; \,\ty{llist} \,=\, \const{LNil} \,\mid\, \const{LCons}(\ty{int},\: \ty{llist})
%\]

%  * consider a natural fragment---datatypes as supported in modern SMT solvers
%    and the SMT-LIB standard, and codatatypes as their duals
%    * mutual recursion, but no polymorphism, nested recursion
%  * integrated with Nelson--Oppen

% * SMT-Lib

%  * one implication is that if $m$ equals $ES(n)$ and $n$ equals $ES(m)$, necessarily
%    $m$ and $n$ must be equal.

%  * perhaps the most commonly used codatatype is that of lazy lists or sequences.
%    using a syntax similar to Standard ML, Haskell, or SMT-LIB

%  * less briefly: codatatypes and why they are useful
%    * in Agda, Coq, Matita -- recently also in Isabelle/HOL \cite{nipkow-et-al-2002}
%    * but not in SMT-LIB 2 \cite{barrett-et-al-2010}

\paragraph{Related Work.} Barrett
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work on datatypes as of 2007, in addition to describing their
implementation in CVC3.
Since then, datatypes have been added not only to CVC4 (a complete rewrite of CVC3)
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} in unpublished work by Leonardo de Moura and to a SPASS-like %prototypical
superposition prover by Wand \cite{wand-2014}.
In his Ph.D.\ thesis \cite{bjoerner-1998}, Bj\o rner introduced a decision procedure for
(co)datatypes in STeP, the Stanford Temporal Prover.
Closely related are the automatic structural
% cruanes-201x,wand-weidenbach-201x,  -- both were rejected
induction in SMT solvers
\cite{reynolds-kuncak-2014}
and superposition provers\cite{kersani-peltier-2013,cruanes-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (\hbox{semi-})decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008} for Isabelle. Its successor, Nitpick
\cite{blanchette-nipkow-2010}, can also generate models involving cyclic
codatatype values. Cyclic values have been studied extensively under the heading
of regular or rational trees---see Carayol and Morvan
\cite{carayol-morvan-2006} and Djellou et al.\ \cite{djelloul-et-al-2008}
for recent work.
The $\mu$-notation is inspired by the $\mu$-calculus
\cite[etc.]{kozen-1983,endrullis-et-al-2011}.

%, notably in the context of constraint logic programming

%  * Ro\c{s}u --- equality of streams is...
%     http://dl.acm.org/citation.cfm?doid=1159803.1159827

%     * point to Barrett et al. for SMT datatypes
%       * about their own work, they say: "our focus is on generality and
%         efficiency rather than immediacy of implementation"
%       * (deal more directly with finite sorts than Barrett et al., Section 6.1)
%     * additional ones since then (e.g. strings?)
%     * anything about codatatypes?
%       * proof assistants like Agda, Coq, etc. have them
%       * Dafny, CoALP
%       * also a lot of theoretical research, some of which is loosely connected,
%         e.g. decision procedure for corecursive functions (Henning in Nijmegen)
%
%     * ODDITY: Oppen 1980: single-constructor, recursive -- infinite values?

@@

%* although nothing prevents composing the decision procedure with theories
%  providing polymorphic types (parametric sorts), such as for arrays (e.g., $\ty{array}(\alpha,\beta)$)

\section{%The Theory of
(Co)datatypes}
\label{sec:co-datatypes}

%    (terminology: freely-generated, inductive, algebraic, ..., sometimes
%    with different meanings; we'll clarify below what we use)

@@

%  * assume for simplicity no indirect recursion, but this does not radically
%    change the argument
%  * assume there are at least two values built with C.
%    they must be different at some point, e.g.
%    C(0, C(1, C(0, ...))) and C(0, C(1, C(1, ...)))
%    can use that to create infinitely alternation patterns
%  * leaves us with the case of a single constructor
% * since we are looking at a specific model, we can assume all uninterpreted
%   types are finite datatypes with nullary constructors corresponding to their
%   elements
% * if two
% * either there is only one ``path'' from ctr to itself
%   or at least two;

\section{The Decision Procedure} % for (Co)datatypes}
\label{sec:the-decision-procedure}

@@

\section{Implementation in CVC4}
\label{sec:implementation-as-a-theory-solver-in-cvc4}

The decision procedure was presented at a high level of abstraction, omitting
quite a few details. This section describes the main aspects of the
implementation within the SMT solver CVC4:\ the integration of the
procedure into CDCL($T$)
\cite{ganzinger-et-al-2004}, the construction of models with $\mu$-terms,
and the extension of the procedure to quantified formulas.
%, and some of the optimizations.

\subsection{A Theory Solver for $\thD$}

The decision procedure is implemented as a \relax{theory solver} of CVC4---that is, a
specialized procedure for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of~$F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant \cite{jovanovic2011sharing}
of the Nelson--Oppen method \cite{nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of
\begin{itemize}
\item the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
\item additional (dis)equalities between variables of types not belonging to
$T_i$.
\end{itemize}
Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

Assume $\Ec$ is initially the set $M_i$ described above. With each equality $t \teq u$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq u$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is assigned an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
The rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For \rn{Split}, we add the exhaustiveness clause %of the form
$\DISC{1} \mathrel\lor \cdots \mathrel\lor \DISC{m}$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

\begin{figure}[t!]
\normalsize
\centering
\flushleft $\rn{Acyclic}(t)$:

\vspace*{-\topsep}

\begin{enumerate}
\item Let $U$ be the datatype equivalence classes of $\Ec$.
\item Repeat until $U \neq \emptyset$:
\begin{itemize}
\item[\ ] For some $\ec{t} \in U$, $\rn{Traverse}( \ec{t}, U, \emptyset )$.
\end{itemize}
\end{enumerate}

\flushleft $\rn{Traverse}( \ec{t}, U, P )$:

\vspace*{-\topsep}

\begin{enumerate}
\item If $\ec{t} \in P$, $\rn{Acyclic}(t)$.
\item If $\ec{t} \in U$:
\begin{enumerate}
\item[2.1.] If $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$, then for each $j = 1, \ldots k$:
\begin{itemize}
\item[2.1.1.] $\rn{Traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} )$.
\end{itemize}
\item[2.2.] $U := U \textbackslash \{ \ec{t} \}$.
\end{enumerate}
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Acyclic}}
\label{fig:acyclic-algo}
\end{figure}

\begin{figure}[t!]
\normalsize
\centering

\flushleft $\rn{Unique}(t, u)$:

\vspace*{-\topsep}

\begin{enumerate}
\item Let $\mathcalx{U}$ be a partition of the equivalence classes of $\Ec$ such that $\ec{t} =_{\mathcalx{U}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
  $t$, $u$ have type $\delta \in \Codata$,
  $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$,
  and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$.

\item Repeat until $\mathcalx{U}$ is unchanged:
\begin{enumerate}
\item[2.1.] $\mathcalx{U}' := \emptyset$.
\item[2.2.] For each $U_i \in \mathcalx{U}$:
\begin{itemize}
\item[2.2.1.] Let $\mathcalx{U}_{\vthinspace i}$ be a partition of $U_i$ such that $\ec{t} =_{\mathcalx{U}_{\vthinspace i}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcalx{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[2.2.2.] $\mathcalx{U}' := \mathcalx{U}' \cup \mathcalx{U}_{\vthinspace i}$.
\end{itemize}
\item[2.3.] $\mathcalx{U} := \mathcalx{U}'$.
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcalx{U}$, $\rn{Unique}(t,u)$.
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, it relies on the algorithms in Figs.~\ref{fig:acyclic-algo} and~\ref{fig:uniqueness-algo}
for determining whether the rules \rn{Acyclic} and \rn{Unique} apply to the set $\Ec$.

For \rn{Acyclic}, Fig.~\ref{fig:acyclic-algo} considers the set $U$ of all datatype equivalence classes of $\Ec$.
We choose an arbitrary $\ec{t}$ in $U$ and call the recursive subprocedure $\rn{Traverse}$,
which takes as input the current equivalence class we are processing, the set of equivalence classes $U$ we have yet to process,
and the set of equivalence classes $P$ we are currently processing.
If a call to $\rn{Traverse}( \ec{t}, U, P )$ is such that $\ec{t} \in P$,
we know that the rule \rn{Acyclic} applies to $t$.
Otherwise, if we have yet to process $\ec{t}$, as indicated by the condition $\ec{t} \in U$,
then if $\ec{t}$ contains a constructor term $\const{C}( t_1, \ldots, t_k )$, 
then we recursively call $\rn{Traverse}$ on each of $\ec{t_1}, \ldots, \ec{t_n}$.
If this succeeds, we remove $\ec{t}$ from $U$.

For \rn{Unique}, Fig.~\ref{fig:uniqueness-algo} considers a partition $\mathcalx{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcalx{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcalx{U}$.
We then refine $\mathcalx{U}$ by 
constructing a partition $\mathcalx{U}_{\vthinspace i}$ of each $U_i \in \mathcalx{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcalx{U}_{\vthinspace i}$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcalx{U}$ for each $j=1, \ldots, k$.
Let $\mathcalx{U'}$ be the union of each of these partitions.
We update $\mathcalx{U}$ to $\mathcalx{U'}$.
This refinement is repeated until $\mathcalx{U}$ is left unchanged.
If the resulting partition $\mathcalx{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
This algorithm is analogous to Hopcroft's algorithm for minimizing
deterministic finite automata \cite{hopcroft-1971}---terms correspond to
states, argument positions $1,\ldots,k$ correspond to input symbols, and the top constructor symbol
of a term generalizes the accepting/rejecting status of a state.

%Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
%To detect whether \rn{Acyclic} applies,
%the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
%It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
%If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
%To recognize when the premise of \rn{Unique} holds,
%the procedure considers the set $\SSS$ of all codatatype equivalence classes. It
%simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
%while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
%and the equivalence of their arguments of ordinary types.
%It then partitions each set %in this partition %in the same manner
%recursively.
%If %after doing so,
%the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
%and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.

The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
Discriminators are part of the signature and not abbreviations.
This requires extending the decision procedure with several
rules, % \cite{barrett-et-al-2007},
which apply uniformly to datatypes and codatatypes.
This approach often leads to better performance because it introduces terms
less eagerly to $\tEc$.
Selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, the solver directly adds
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality.
To reduce the number of unique literals considered by the calculus, we
compute a normal form for literals as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is smaller than $u$ with respect to some term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.

% AJR : this paragraph may have been what confused Reviewer #2. It does suggest that we are copying verbatim what was in CVC3 (which is what Reviewer #2 thought).
% Since we aren't saying anything new beyond what is said in Barrett et al (and honestly, we didn't critically evaluate the performance of the updated implementation), 
% maybe we should just remove this paragraph.

% JB to AJR: I'd rather keep the para, based on feedback by Pascal.

As Barrett et al.\ %\cite{barrett-et-al-2007}
observed for their procedure,
it is both theoretically and
empirically beneficial to delay applications of \rn{Split} as long as
possible. Similarly, \rn{Acyclic} and \rn{Unique} are fairly expensive because
they require traversing the equivalence classes, which is why they are part of
phase~2.

\subsection{Model Construction}

When instructed to do so, the implementation produces models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we %may
construct a map $\ValC$ from 
the equivalence classes of $\Ec$ to closed normal $\MU$-terms.
Recall that our construction of $\ValC$ requires choosing values $\MU {\Varec{x}}.\; t$ for each $\Varec{x} \in \FV( \ValC )$
such that $\MU {\Varec{x}}.\; t$ does not occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To choose values, % $\MU {\Varec{x}}.\; t$,
we use a fair enumerator for each (co)datatype $\tau$,
which lists all the values
of type $\tau$ in a normal interpretation. %, which are closed normal $\MU$-terms.

%Concretely,
For codatatypes, we enumerate a stream of \relax{every} $\MU$-term of our signature $\Sig$ in a fair manner,
discarding those that have free variables or are not normal.
For those that are closed and normal, we then check whether they occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
This set does not need to be explicitly constructed.
%We do not explicitly construct the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To determine if a term $\MU {\Varec{x}}.\; t$ occurs in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$, 
we check if it is $\alpha$-equivalent to a closed term $(\MU {\Varec{y}}.\; u) \{ \Varec{x} \mapsto z \}$ for
some $\MU {\Varec{y}}.\; u$ in the range of $\ValC$ and variable $z$.
This can be efficiently achieved by matching the term $\MU {\Varec{x}}.\; t$ with %the term
$\MU {\Varec{y}}.\; u$.

%Once the mapping $\ValC$ is constructed, our model $\J$ interprets
%$\J( x ) = \ValC \ec{x}$ for each variable $x$ occurring free in $\Ec$.

\subsection{Extension to Quantified Formulas}

While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete 
instantiation-based 
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to quantified
formulas involving datatypes and codatatypes.

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas can entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
By assuming that uninterpreted types have infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.

Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the codatatype $\ty{stream}_{\vthinspace\ty{u}}$ %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$,
which states that either the cardinality of $\ty{u}$ is greater than one or
$s$ must be equal to $t$.

%\section{Examples}
%\label{sec:examples}

\section{Evaluation on Isabelle Problems}
\label{sec:evaluation-on-isabelle-problems}

@@

\section{Conclusion}
\label{sec:conclusion}

@@

%Finally, it might be possible to go
%further in terms of supporting nested and mixed (co)recursion and quantified
%formulas over (co)datatypes in solvers.

>>>>

@@

\section{Introduction -- JB 2 columns}
\label{sec:introduction}

%  * a general introduction to automatic theorem proving \\

%  * relation to AI
%   * among the earliest AI goals
%   * yet somewhat disconnected from mainstream AI
%     (with some exceptions, e.g. Urban ..., MaSh)
%
%  * followed by an intro to SMT and what it's used for \\
%  * followed to why theories are important (more efficient reasoning) \\
%  * followed to why (co)datatypes are important
%
%  * both for proving
%   * not finitely axiomatizable
%  * and for model finding
%   * Nunchaku

@@

@@

\section{%The Theory of
(Co)datatypes -- AJR 0.75 column}
\label{sec:co-datatypes}

@@

\section{The Decision Procedure -- AJR 4 columns} % for (Co)datatypes}
\label{sec:the-decision-procedure}

@@

\section{Evaluation -- JB 1 column} % on Isabelle Problems}
\label{sec:evaluation}

@@

\section{Conclusion -- JB 0.25 column}
\label{sec:conclusion}

@@
